---
title: "Is it Really the Economy?"
date: "`r Sys.Date()`"
author: "Michelle Liu"
output:
  pdf_document:
    toc: no
    toc_depth: '2'
editor_options: 
  markdown: 
    wrap: 72
---
#Project Details
#Project analyzing whether the economy impacts electoral results
#data: ANES data about electoral results by party by state, state-level average disposable income, Census data about demographics


#Set Up
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning = FALSE, 
                      fig.width = 16/2, fig.height = 9/2)
```

#Packages
```{r packages, include=F}
library(ggplot2)
library(dplyr)
library(knitr)
library(tidyverse)
library(readxl)
library(priceR)
```

## Read data
```{r}
rm(list=ls())

setwd('/Users/michelleliu/Documents/UNC/Fall 2023/POLI 381/Final Project/Data')

#import data
rdi <- read.csv("rdi.csv")

#data cleaning
rdi = rdi %>%
  select(state, year, RDI) %>%
  filter(year %in% c(1972, 1976, 1980, 1984, 1988, 1992, 
                     1996, 2000, 2004, 2008, 2012, 2016, 2020)) %>%
  filter(!is.na(state)) %>%
  mutate(state = factor(state, levels = c(state.name, 'District of Columbia'),
                                         labels = c(state.abb, 'DC')))

rdi_1972 <- rdi %>%
  filter(year == 1972) 
rdi_1972$vote_dem <- NA
rdi_1972$vote_rep <- NA

#We're using the data set with vote_dem and vote_rep
pres <- read_csv("1976-2020-president.csv") %>%
  filter(party_simplified %in% c('DEMOCRAT', "REPUBLICAN")) %>%
  mutate(vote_share = candidatevotes/totalvotes) %>%
  group_by(year, state=state_po, party_simplified) %>%
  summarise(vote_share = sum(vote_share)) %>% 
  # making a wide dataset with columns for Dem and Rep vote shares
  pivot_wider(names_from = party_simplified, values_from = vote_share) %>%
  rename(vote_dem = DEMOCRAT, vote_rep = REPUBLICAN)

#Merge
data <- left_join(pres, rdi, by = c("state", "year"))
data <- rbind(data, rdi_1972)
```

#Account for inflation using priceR
```{r}
#data about average disposable income was not adjusted for inflation
inflation_dataframe = retrieve_inflation_data('US')
countries_dataframe = show_countries()
data = data %>%
  mutate(RDI = adjust_for_inflation(RDI, year, 'US', to_date = 2020,
                                   inflation_dataframe = inflation_dataframe,
                                   countries_dataframe = countries_dataframe))

rm(countries_dataframe)
rm(inflation_dataframe)
```

#Find RDI change rate using lag() function
#https://dplyr.tidyverse.org/reference/lead-lag.html
```{r}
data <- data %>%
  arrange(state, year) %>%
  group_by(state) %>%
  mutate(lag_rdi = lag(RDI)) %>%
  mutate(RDI_rate = (RDI-lag_rdi)/lag_rdi) %>%
  filter(year != 1972)
```

#Account for incumbency
```{r}
data$incumbent <- NA

#when Democrats are incumbent
data$incumbent[data$year %in% c(1980, 1996, 2000, 2012, 2016)] <- 1

#When Republicans are incumbent
data$incumbent[data$year %in% c(1976, 1984, 1988, 1992, 2004, 2008, 2020)] <- 0

#Find incumbent vote share
data <- data %>%
  mutate(vote = ifelse(incumbent==1, vote_dem, vote_rep))

```
    
#Add unemployment to data   
#Using unemployment as a robustness check for average disposable income
```{r}
data$year <- as.integer(data$year)

state_unem <- read_excel("/Users/michelleliu/Documents/UNC/Fall 2023/POLI 381/Final Project/Data/staadata_cleaned.xlsx") %>%
  filter(!is.na(Employment)) %>%
  select(-c(1, 4:9)) %>%
  rename(state = "...2", year = "...3",rate  = "...10") %>%
  filter(state != c("Los Angeles County", "New York city")) %>%
  mutate(state = factor(state, levels = c(state.name,"District of Columbia"), 
                        labels = c(state.abb, "DC"))) %>%
  mutate(year = as.integer(year))

data <- left_join(data, state_unem, by = c("state", "year"))
    
rm(state_unem)
```

#using the African American population % as a control variable
#getting this data from the Census
black for census years, based on column names in the IPUMS codebook
```{r}
#1970
c1970 <- read.csv("/Users/michelleliu/Documents/UNC/Fall 2023/POLI 381/Final Project/Data/1970_state.csv") %>%
  select(YEAR, STATE, CBW001, CBW002, CBW003, CBW004, CBW005) %>%
  mutate(black = CBW002 / rowSums(
    select(., CBW001, CBW002, CBW003, CBW004, CBW005))) %>%
  mutate(STATE = factor(STATE, levels = c(state.name, "District Of Columbia"), 
                        labels = c(state.abb, "DC"))) %>%
  select(YEAR, STATE, black)

#1980
c1980 <- read.csv("/Users/michelleliu/Documents/UNC/Fall 2023/POLI 381/Final Project/Data/1980_state.csv") %>%
  select(YEAR, STATE, C6X001, C6X002, C6X003, C6X004, C6X005) %>%
  mutate(black = C6X002 / rowSums(
    select(., C6X001, C6X002, C6X003, C6X004, C6X005))) %>%
           mutate(STATE = factor(STATE, 
                                 levels = c(state.name, "District Of Columbia"), 
                                 labels = c(state.abb, "DC"))) %>%
  select(YEAR, STATE, black)                

#1990
c1990 <- read.csv("/Users/michelleliu/Documents/UNC/Fall 2023/POLI 381/Final Project/Data/1990_state.csv") %>%
  select(YEAR, STUSAB, EUY001, EUY002, EUY003, EUY004, EUY005) %>%
  rename(STATE = STUSAB) %>%
  mutate(black = EUY002 / rowSums(
    select(., EUY001, EUY002, EUY003, EUY004, EUY005))) %>%
  select(YEAR, STATE, black)

#2000
c2000 <- read.csv("/Users/michelleliu/Documents/UNC/Fall 2023/POLI 381/Final Project/Data/2000_state.csv") %>%
  select(YEAR, STATE, FMR001, FMR002, FMR003, FMR004, FMR005, FMR006, FMR007) %>%
  mutate(STATE = factor(STATE, levels = c(state.name, "District Of Columbia"), 
                        labels = c(state.abb, "DC"))) %>%
  mutate(black = FMR002 / rowSums(
    select(., FMR001, FMR002, FMR003, FMR004, FMR005, FMR006, FMR007))) %>% 
  select(YEAR, STATE, black)

#2010
c2010 <- read.csv("/Users/michelleliu/Documents/UNC/Fall 2023/POLI 381/Final Project/Data/2010_state.csv") %>%
  filter(STUSAB != "PR") %>%
  select(YEAR, STUSAB, H7X001, H7X002, H7X003, H7X004, 
         H7X005, H7X006, H7X007, H7X008) %>%
  rename(STATE = STUSAB) %>%
  mutate(black = H7X003 / H7X001) %>%
  select(YEAR, STATE, black)

#2020
c2020 <- read.csv("/Users/michelleliu/Documents/UNC/Fall 2023/POLI 381/Final Project/Data/2020_state.csv") %>%
  filter(STUSAB != "PR") %>%
  select(YEAR, STUSAB, U7J001, U7J002, U7J003, U7J004, 
         U7J005, U7J006, U7J007, U7J008) %>%
  rename(STATE = STUSAB) %>%
  mutate(black = U7J003 / U7J001) %>%
  select(YEAR, STATE, black)

```    
  

Find the number of African Americans in each state for non-Census years 
Find slope in black between election years
Add part of the slope to the black of the previous election year
```{r}
c1976 <- c1970 %>%
  left_join(c1980, by = "STATE") %>%
  mutate(YEAR=1976, black = black.x + 
           (black.y - black.x) * (6 / 10)) %>%
  select(YEAR, STATE, black)

c1984 <- c1980 %>%
  left_join(c1990, by = "STATE") %>%
  mutate(YEAR=1984, black = black.x + 
           (black.y - black.x) * (4 / 10)) %>%
  select(YEAR, STATE, black)

c1988 <- c1980 %>%
  left_join(c1990, by = "STATE") %>%
  mutate(YEAR=1988, black = black.x + 
           (black.y - black.x) * (8 / 10)) %>%
  select(YEAR, STATE, black)

c1992 <- c1990 %>%
  left_join(c2000, by = "STATE") %>%
  mutate(YEAR=1992, black = black.x + 
           (black.y - black.x) * (2 / 10)) %>%
  select(YEAR, STATE, black)

c1996 <- c1990 %>%
  left_join(c2000, by = "STATE") %>%
  mutate(YEAR=1996, black = black.x + 
           (black.y - black.x) * (6 / 10)) %>%
  select(YEAR, STATE, black)

c2004 <- c2000 %>%
  left_join(c2010, by = "STATE") %>%
  mutate(YEAR=2004, black = black.x + 
           (black.y - black.x) * (4 / 10)) %>%
  select(YEAR, STATE, black)

c2008 <- c2000 %>%
  left_join(c2010, by = "STATE") %>%
  mutate(YEAR=2008, black = black.x + 
           (black.y - black.x) * (8 / 10)) %>%
  select(YEAR, STATE, black)

c2012 <- c2010 %>%
  left_join(c2020, by = "STATE") %>%
  mutate(YEAR=2012, black = black.x + 
           (black.y - black.x) * (2 / 10)) %>%
  select(YEAR, STATE, black)

c2016 <- c2010 %>%
  left_join(c2020, by = "STATE") %>%
  mutate(YEAR=2016, black = black.x + 
           (black.y - black.x) * (6 / 10)) %>%
  select(YEAR, STATE, black)

```

Rbind the data about demographics to the main data frame
```{r}
black_df <- rbind(c1976, c1980, c1984, c1988, c1992, c1996, 
                       c2000, c2004,c2008, c2012, c2016, c2020)

black_df <- black_df %>%
  rename(year = YEAR, state = STATE)

data <- left_join(data, black_df, by = c("state", "year"))

rm(c1970)
rm(c1976)
rm(c1984)
rm(c1988)
rm(c1990)
rm(c1980)
rm(c1992)
rm(c1996)
rm(c2004)
rm(c2008)
rm(c2000)
rm(c2010)
rm(c2012)
rm(c2016)
rm(c2020)
rm(black_df)
rm(cities)
rm(pres)
rm(rdi)
rm(rdi_1972)
```

Recoding so we interpret as percentage point change
```{r}
data$black <- data$black*100
data$vote = data$vote*100
data$RDI_rate <- data$RDI_rate*100
```

Histograms of variables used for regression
```{r}
hist(data$black, xlab = '% of Population that is Black', 
     ylab = 'Frequency', 
     main = 'Distribution of Black % of Population In Each State', 
     xlim=c(0,80))

hist(data$vote, xlab = 'Distribution of Incumbent Presidential Vote Share In Each State', 
     ylab = 'Frequency', 
     main = 'Incumbent Presidential Vote Share')

hist(data$RDI_rate, main = "Distribution of RDI % Change In Each State", 
     xlab = "RDI % change, adjusted for inflation")

hist(data$rate, xlab = 'Distribution of Unemployment Rate In Each State',
     ylab = 'Frequency', 
     main = 'Unemployment Rate')

```

#DC is skewing demographics, so remove DC
```{r}
data <- data[data$state!= "DC",]

hist(data$black, xlab = '% of Population That Is Black', 
     ylab = 'Frequency', 
     main = 'Distribution of Black % of Population In Each State')
```

Analyzing covariance to see what should be added as a control variable in the later regressions
```{r}
cor(data$RDI_rate, data$black)
cor(data$vote, data$black)
```

Bivariate regression on the overall data
Plot scatterplot of regression
Residual analysis: plot residuals histogram, residual scatterplots
```{r}
fit1 <- lm(vote ~ RDI_rate, data = data)
summary(fit1)

#scatterplot with regression line
ggplot(data, aes(x = RDI_rate, y = vote)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "Relationship Btwn. RDI % Change and Incumbent Vote Share",
       x = "RDI % Change, adjusted for inflation",
       y = "Incumbent Presidential Vote Share")

#evaluate residuals for abnormal patterns
resid <- fit1$residuals
fv <- fit1$fitted.values
resid_fv <- as.data.frame(cbind(resid, fv))

hist(resid, xlab = 'Residuals', ylab = 'Frequency', main = 'Distribution of Residuals from Bivariate Regressionn')

ggplot(data=data, aes(x=RDI_rate, y = resid)) + 
  geom_point() + 
  labs(title = "RDI % Change Plotted Against Residuals in Bivariate Regression",
       x = "RDI % Change, adjusted for inflation",
       y = "Residuals")

ggplot(data=resid_fv, aes(x=fv, y = resid)) + 
  geom_point() + 
  labs(title = "Bivariate Fitted Values Plotted Against Residuals",
       x = "Fitted Values",
       y = "Residuals")
```

Multivariate regression on overall data
```{r}
fit2 <- lm(vote ~ RDI_rate + black, data = data)
summary(fit2)

resid2 <- fit2$residuals
fv2 <- fit2$fitted.values
resid_fv2 <- as.data.frame(cbind(resid2, fv2))

hist(resid2, xlab = 'Residuals', ylab = 'Frequency', 
     main = 'Distribution of Residuals from Multivariate Regressionn')

ggplot(data=resid_fv2, aes(x=fv2, y = resid2)) + 
  geom_point() + 
  labs(title = "Multivariate Fitted Values Plotted Against Residuals",
       x = "Fitted Values",
       y = "Residuals")
```
    
Interaction regression on overall data with visualizing residuals
```{r}
fit3 <- lm(vote ~  RDI_rate + black + 
             RDI_rate:black, data = data)
summary(fit3)

resid3 <- fit3$residuals
fv3 <- fit3$fitted.values
resid_fv3 <- as.data.frame(cbind(resid3, fv2))

hist(resid3, xlab = 'Residuals', ylab = 'Frequency', 
     main = 'Distribution of Residuals from Interaction Regressionn')

ggplot(data=resid_fv3, aes(x=fv3, y = resid3)) + 
  geom_point() + 
  labs(title = "Interaction Regression Fitted Values Plotted Against Residuals",
       x = "Fitted Values",
       y = "Residuals")
```

Multivariate and interaction regressions by year
```{r}
elections <- c(1976, 1980, 1984, 1988, 1992, 1996, 
               2000, 2004, 2008, 2012, 2016, 2020)

for (i in elections) {
  one_year <- filter(data, year == i)
  cat(i)
  print(summary(lm(vote ~ RDI_rate + 
                     black + RDI_rate:black, 
                   data = one_year)))}

rm(one_year)
```

Robustness testing
```{r}
robust <- lm(vote ~ rate + black + black:rate, data = data)
r_summary <- summary(robust)
r_summary
```
    
